Trait plasticity and covariance along a continuous soil moisture gradient

### 2020-02-17 FVT Manuscript/manuscript.tex
commit

\abstract{
Water is perhaps the greatest environmental determinant of plant yield and fitness, yet our understanding of plant-water relations is limited because it is primarily informed by experiments that have treated soil moisture variability as two simple discrete levels - wet and dry - rather than a continually varying environmental gradient. Here we used experimental and statistical methods based on function-valued traits to study responses to continuously varying soil moisture gradient in a suite of physiological and morphological traits in two species and multiple genotypes of the model grass \textit{Brachypodium}. We find that the majority of traits exhibit non-linear responses to soil moisture variability. We also observe differences in the shape of these non-linear responses between traits, species, and genotypes. Emergent phenomena arise from this variation including changes in trait correlations and evolutionary constraints as a function of soil moisture. These results point to the importance of considering non-linearity in plant-water relations to predict outcomes of adaptation to changing climates.

@@

A deep body of literature aims to understand the traits and strategies used by plants to acclimate to variable soil water content. Quantitative genetic variation in myriad drought resistance traits have been observed in natural populations and laboratory model systems. In particular, natural populations of Brassicaceae species (including \emph{Arabidopsis thaliana}) and \emph{Brachypodium} harbor variation both constitutive and inducible traits mediating plant-water relations, including water use efficiency (Des Marais \emph{et al.} 2012, 2017; Edwards \emph{et al.} 2012; Greenham \emph{et al.} 2017), production of compatible osmolytes (Kesari \emph{et al.} 2012; Des Marais \emph{et al.} 2017), leaf anatomy (Skirycz \emph{et al.} 2011; Verelst \emph{et al.} 2013; Dittberner \emph{et al.} 2018), root-shoot biomass partitioning (Des Marais \emph{et al.} 2012, 2017), and many others (Verslues and Juenger 2011; Edwards \emph{et al.} 2012; Juenger 2013; Luo \emph{et al.} 2016; Yarkhunova \emph{et al.} 2016; Lenk \emph{et al.} 2019). Among the numerous plant traits that can and have been measured, few have been studied more extensively than specific leaf area (SLA - often reported as it's inverse, leaf mass per area or LMA). SLA provides a description of leaf architecture that is central to the leaf economics spectrum, a theory which seeks to explain variation in leaf physiological strategies, from more conservative (low SLA) to more productive (high SLA) (Wright \emph{et al.} 2004). In the context of drought stress, it is intuitive that low SLA might be adaptive as lower leaf surface area is expected to reduce water loss through transpiration. In several cases, reductions in SLA have been reported under drought conditions (Casper \emph{et al.} 2001).

@@

To characterize phenotypic responses to our simulated drought, we measured a suite of developmental and physiological traits. Plants were harvested in five cohorts over five days. Each day, half of the sampled plants were harvested for above and below ground biomass, total above ground green area, \(\delta C_{13}\), N content, C content. The other half were harvested and assessed for specific leaf area (SLA) and relative water content (RWC). Above ground biomass was measured after drying leaf material overnight at 60°C. Below ground biomass was measured after washing the soil matrix from roots and drying them overnight at 60°C. Above ground leaf area was estimated by laying plants flat between plates of clear plexiglass and imaging with a Nikon 5300 digital camera at a fixed distance with a 35mm Nikkor lens. Total green pixels were counted for each image with Easy Leaf Area with settings shown in Figure \ref{fig:leafarea}. SLA was calculated by scanning the two youngest fully emerged leaves. Leaf area was calculated from these same images using Easy Leaf Area. Above and belowground biomass was measured after leaves were dried and SLA was calculated as \(leaf\ area / biomass_{dry}\). Leaf tissues for \(\delta C_{13}\), \(\delta N_{15}\), nitrogen (hereafter \enquote{N}) content, and carbon (hereafter \enquote{C}) content were ground to a fine powder and processed by the UC Davis Stable Isotope Facility. These leaves were also used to calculate RWC. Prior to drying, fresh leaves were weighed (\(biomass_{fresh}\)) and then submerged under water in 15mL falcon tubes for several hours. They were then weighed (\(biomass_{turgid})\), oven-dried overnight, and weighed again (\(biomass_{dry}\)). RWC was calculated as \((biomass_{fresh}-biomass_{dry})/(biomass_{turgid}-biomass_{dry})\)\}.

@@

For the purposes of modeling phenotypic responses to variation in soil moisture content, we considered soil moisture content as the final soil moisture on day 14 of the dry down period for each plant, referred to in figures as \(Soil\  moisture (\%_{final})\). A major challenge in studying function-valued traits is model selection. That is, identifying the functions that best describe the curvature (or lack thereof) in the shape of phenotypic responses to environmental gradients. Quadratic and natural splines have been suggested as potential functions to model non-linearities (Meyer 2005), but selecting the appropriate function is challenging. Akaike information criterion (AIC) selection based on contrasting multiple complex models offers an effective means to balance predictability with over-fitting (Griswold \emph{et al.} 2008; Gomulkiewicz \emph{et al.} 2018). Thus, we began with the complex model for traits as described below.

@@

We calculated several statistics summarizing evolutionary constraint as described in Kirkpatrick (2009). First, for each species, we estimated the \(G\) matrix of genetic covariances between genotype trait means at different levels of soil moisture. We then calculated, using the \emph{prcomp} function in R, the eigenvalues of each mean standardized (trait values divided by mean) \(G\) matrix, \(\lambda_i\). From these we then calculated the \emph{number of effective dimensions}, \(n_{D}\), equal to the sum of the eigenvalues divided by the largest eigenvalue:

@@

@@

We evaluated the degree to which traits show linear or non-linear shapes using an AIC model selection approach from a full model which included quadratic and natural spline parameters relating soil moisture content to plant phenotypes. We observed significant (\(\alpha < 0.05\)) non-linear components (quadratic, spline, or both) in the final models for all of the traits which included an environmental (water content) predictor (Table \ref{tab:models}, Figure \ref{fig:curves}) except N content in \emph{B. distachyon}, C content in \emph{B. sylvaticum}, and shoot mass in \emph{B. sylvaticum} where non-linear environmental predictors were included in the final models chosen by AIC but did not significantly explain trait variance (p \textgreater{} 0.05). In \emph{B. distachyon}, all of the traits included at least one non-linear environmental predictor. In contrast, SLA, total biomass and shoot mass were not predicted by environment in \emph{B. sylvaticum}. Interestingly, of all the traits which were predicted by environment in both \emph{B. distachyon} and \emph{B. sylvaticum}, the shape -- whether a quadratic versus a spline function was included in the model -- was different in both species with the exception of carbon to nitrogen ratio.

@@

\begin{figure*}[!h]
\includegraphics[width=\textwidth]{../Figures/pca_byspecies} \caption{\textbf{Plasticity through multivariate trait space.} (A) Plasticity across all traits was calculated as distance between scaled phenotype for each genotype between different levels of soil moisture. Box plots indicate species median and 25th and 75th percentiles with whiskers extending to 1.5 times the interquartile range. "*" indicates significant differences between species (t-test, \(\alpha < 0.05\)). Principal component analysis of scaled phenotypic responses to soil moisture gradient among genotypes of both species. Upper panels show genotype means across soil mositure content. Percent values in axis titles indicate percent variance explained by that principal component. Lower panels show eigenvectors of each trait. (B) PC1 and PC2. (C) PC3 and PC4.}\label{fig:pca}
\end{figure*}

@@

We also tested whether there was significant natural variation for the traits measured between genotypes in each species by looking at the parameters in the final model for each trait. Interestingly, in both species and for all traits except relative water content in \emph{B. sylvaticum} the final model included a significant genotype term, indicating significant differences between genotypes in magnitude of traits across all levels of soil moisture (Table \ref{tab:models}, Figure \ref{fig:curves}). For other traits there are clear distinctions between the two species. For example, \(\delta C_{13}\) was considerably higher in \emph{B. sylvaticum} (Figure \ref{fig:curves}). For SLA, while \emph{B. distachyon} showed a strong response to soil moisture, especially under the driest conditions, SLA in \emph{B. sylvaticum} was not responsive to soil moisture. In contrast, \emph{B. sylvaticum} appears to show a more dramatic response in leaf composition estimated by N content and C:N ratio.

@@

Significant interactions between genotype and environmental parameters in a final model indicate the presence of genetic variation for plasticity (GxE) (Via and Lande 1985). For those GxE interactions where the environmental parameter is non-linear, significant GxE indicates genetic variation for the shape of reaction norms. SLA showed a significant interaction between genotype and soil moisture in \emph{B. distachyon}. In \emph{B. sylvaticum}, nitrogen content, carbon to nitrogen ratio, above ground green area, and root to shoot ratio all showed significant interactions between genotype and soil moisture (Table \ref{tab:models}, Figure \ref{fig:curves}). In each of these cases, the interaction between genotype and environment involved a non-linear environmental predictor, indicating not only variation for the magnitude of plasticity (i.e.~slope) but also variation in the shape of responses.

@@

We found significant non-linearity in response to a soil moisture gradient for all measured traits in at least one of the two species sampled. The best-fit function for some traits was quadratic, while other traits showed more complex responses to the environment which were best fit by a spline function. These results offer new insights with respect to the study of plant response to soil drying. By focusing on the curvature of phenotypic response as the explicitly modeled trait, we avoid contrasts of trait values expressed at arbitrary levels of soil water content which may obscure different thresholds of response among the diverse genotypes under study. SLA in \emph{B. distachyon} exhibits this pattern, as two accessions show a threshold-like response in decreasing SLA as soils become drier, and three accessions express their highest SLA at intermediate SWC. Leaf N content (on a leaf-mass basis) in \emph{B. sylvaticum} likewise shows considerable diversity of response with two accessions expressing their lowest values at intermediate SWC, one accession expressing its highest Leaf N at intermediate SWC, and one accession showing a nearly linear response along the SWC gradient.

@@

From a practical perspective this work highlights the value of function-valued trait approaches that may be extended to studying plant-water relations in agricultural settings. In this experiment, we investigated variation in plant responses to a gradient of soil moisture using six watering levels, which in combination with random variation in water capacity of experimental pots, produced a continuous gradient of soil moisture ranging from field capacity of the soil to strongly water-limited. In the field, multiple watering regimes in combination with random variability between plots may produce similar gradients of soil moisture. In this experiment, water content was measured gravimetrically. New sensing technologies may be useful for quantifying soil moisture in an analogous fashion, to define soil moisture quantitatively and then apply function-valued statistical approaches to contrast trait expression among genotypes.

\hypertarget{concluding-remarks}{%
\subsubsection{Concluding remarks}\label{concluding-remarks}}

Describing traits and their complex relationships with environmental and genetic variation continue to be challenges for biology. This goal is worthy of further effort if we are to address the challenges posed by climate change and understand organism evolution. Fortunately, new approaches such as analyses of function-valued traits and multivariate trait covariance analyses facilitate new discoveries. We hope that this study, by demonstrating the use of these approaches to yield novel insight into the influence of environment on phenotypes, trait correlations, and evolutionary constraint, will inspire further work to tackle the challenge of studying organisms and environments.

@@

>>>>

\abstract{
Water is perhaps the greatest environmental determinant of plant yield and fitness, yet our understanding of plant-water relations is limited because it is primarily informed by experiments that have treated soil moisture variability as two simple discrete levels - wet and dry - rather than a continually varying environmental gradient. Here we used experimental and statistical methods based on function-valued traits to study responses to continuously varying soil moisture gradient in a suite of physiological and morphological traits in two species and multiple genotypes of the model grass \textit{Brachypodium}. We find that the majority of traits exhibit non-linear responses to soil moisture variability. We also observe differences in the shape of these non-linear responses between traits, species, and genotypes. Emergent phenomena arise from this variation including changes in trait correlations and evolutionary constraints as a function of soil moisture. These results point to the importance of considering non-linearity in plant-water relations to predict plastic and evolutionary responses to changing climates.

@@

A deep body of literature aims to understand the traits and strategies used by plants to acclimate to variable soil water content. Quantitative genetic variation in myriad drought resistance traits have been observed in natural populations and laboratory model systems. In particular, natural populations of Brassicaceae species (including \emph{Arabidopsis thaliana}) and \emph{Brachypodium} harbor variation both constitutive and inducible traits mediating plant-water relations, including water use efficiency (Des Marais \emph{et al.} 2012, 2017; Edwards \emph{et al.} 2012; Greenham \emph{et al.} 2017), leaf chemistry (Kesari \emph{et al.} 2012; Des Marais \emph{et al.} 2017), leaf anatomy (Skirycz \emph{et al.} 2011; Verelst \emph{et al.} 2013; Dittberner \emph{et al.} 2018), root-shoot biomass partitioning (Des Marais \emph{et al.} 2012, 2017), and many others (Verslues and Juenger 2011; Edwards \emph{et al.} 2012; Juenger 2013; Luo \emph{et al.} 2016; Yarkhunova \emph{et al.} 2016; Lenk \emph{et al.} 2019). Among the numerous plant traits that can and have been measured, few have been studied more extensively than specific leaf area (SLA - often reported as it's inverse, leaf mass per area or LMA). SLA provides a description of leaf architecture that is central to the leaf economics spectrum, a theory which seeks to explain variation in leaf physiological strategies, from more conservative (low SLA) to more productive (high SLA) (Wright \emph{et al.} 2004). In the context of drought stress, it is intuitive that low SLA might be adaptive as lower leaf surface area is expected to reduce water loss through transpiration. In several cases, reductions in SLA have been reported under drought conditions (Casper \emph{et al.} 2001). For all of these traits, the shape of plastic responses to variation in water availability remain largely unmeasured.

@@

To characterize phenotypic responses to our experimental soil moisture gradient, we measured a suite of developmental and physiological traits. Plants were harvested in five cohorts over five days. Each day, half of the sampled plants were harvested for above and below ground biomass, total above ground green area, \(\delta C_{13}\), N content, C content. The other half were harvested and assessed for specific leaf area (SLA) and relative water content (RWC). Above ground leaf area was estimated by laying freshly harvested plants flat between plates of clear plexiglass and imaging with a Nikon 5300 digital camera at a fixed distance with a 35mm Nikkor lens. Total green pixels were counted for each image with Easy Leaf Area (\url{https://github.com/heaslon/Easy-Leaf-Area}) with settings shown in Figure \ref{fig:leafarea}. Above ground biomass was measured after drying leaf material overnight at 60°C and then for several weeks at room temperature. Below ground biomass was measured after washing the soil matrix from roots and drying them overnight at 60°C and then for several weeks at room temperature. Above and belowground biomass was measured after leaves and roots were dried. Leaf tissues for \(\delta C_{13}\), \(\delta N_{15}\), nitrogen (hereafter \enquote{N}) content, and carbon (hereafter \enquote{C}) content were ground to a fine powder and processed by the UC Davis Stable Isotope Facility. SLA was calculated by scanning the two youngest fully emerged leaves with a 1 \(cm^2\) red square. Leaf area in \(mm^2\) was calculated from these same images using Easy Leaf Area. SLA was calculated as \(leaf\ area / biomass_{dry}\). These leaves were also used to calculate RWC. Prior to drying, fresh leaves were weighed (\(biomass_{fresh}\)) and then submerged under water in 15mL falcon tubes for several hours. They were then weighed (\(biomass_{turgid})\), oven-dried overnight, and weighed again (\(biomass_{dry}\)). RWC was calculated as \((biomass_{fresh}-biomass_{dry})/(biomass_{turgid}-biomass_{dry})\).

@@

For the purposes of modeling phenotypic responses to variation in soil moisture content, we considered soil moisture content as the final soil moisture on day 14 of the dry down period for each plant, referred to in figures as \(Soil\ moisture (\%_{final})\). A major challenge in studying function-valued traits is model selection. That is, identifying the functions that best describe the curvature (or lack thereof) in the shape of phenotypic responses to environmental gradients. Quadratic and natural splines have been suggested as potential functions to model non-linearities (Meyer 2005), but assuming the appropriate function is problematic. Akaike information criterion (AIC) selection based on contrasting multiple complex models offers an effective means to balance predictability with over-fitting (Griswold \emph{et al.} 2008; Gomulkiewicz \emph{et al.} 2018). Thus, we began with the complex model for traits as described below.

@@

We calculated several statistics summarizing evolutionary constraint as described in Kirkpatrick (2009). First, for each species, we estimated the \(G\) matrix of genetic covariances between mean scaled genotype trait means at different levels of soil moisture. We then calculated, using the \emph{prcomp} function in R, the eigenvalues of each mean standardized (trait values divided by mean) \(G\) matrix, \(\lambda_i\). From these we then calculated the \emph{number of effective dimensions}, \(n_{D}\), equal to the sum of the eigenvalues divided by the largest eigenvalue:

@@

\begin{figure*}[!h]
\includegraphics[width=\textwidth]{../Figures/pca_byspecies} \caption{\textbf{Plasticity through multivariate trait space.} (a) Plasticity across all traits was calculated as distance between scaled phenotype for each genotype between different levels of soil moisture. Box plots indicate species median and 25th and 75th percentiles with whiskers extending to 1.5 times the interquartile range. "*" indicates significant differences between species (t-test, \(\alpha < 0.05\)). Principal component analysis of scaled phenotypic responses to soil moisture gradient among genotypes of both species. Upper panels show genotype means across soil mositure content. Percent values in axis titles indicate percent variance explained by that principal component. Lower panels show eigenvectors of each trait. (b) PC1 and PC2. (c) PC3 and PC4.}\label{fig:pca}
\end{figure*}

@@

We evaluated the degree to which traits show linear or non-linear shapes using an AIC model selection approach from a full model which included quadratic and natural spline parameters relating soil moisture content to plant phenotypes. We observed significant (\(\alpha < 0.05\)) non-linear components (quadratic, spline, or both) in the final models for all of the traits which included an environmental (water content) predictor (Table \ref{tab:models}, Figure \ref{fig:curves}) except N content in \emph{B. distachyon}, C content in \emph{B. sylvaticum}, and shoot mass in \emph{B. sylvaticum} where non-linear environmental predictors were included in the final models chosen by AIC but did not significantly explain trait variance (p \textgreater{} 0.05). In \emph{B. distachyon}, all of the traits included at least one non-linear environmental predictor. In contrast, SLA, total biomass and shoot mass did not include environmental predcitor in selected models in \emph{B. sylvaticum}. Interestingly, of all the traits which were predicted by environment in both \emph{B. distachyon} and \emph{B. sylvaticum}, the shape -- whether a quadratic versus a spline function was included in the model -- differed between the two species with the exception of carbon to nitrogen ratio.

@@

@@

We also tested whether there was significant natural variation for the traits measured between genotypes in each species by looking at the parameters in the final model for each trait. Interestingly, in across species, most traits (20/24) included a significant genotype term in the final model, indicating significant differences between genotypes in magnitude of traits across all levels of soil moisture (Table \ref{tab:models}, Figure \ref{fig:curves}). Though not formally tested, for other traits there are clear distinctions between the two species. For example, \(\delta C_{13}\) was considerably higher in \emph{B. sylvaticum} (Figure \ref{fig:curves}). For SLA, while \emph{B. distachyon} showed a strong response to soil moisture, especially under the driest conditions, SLA in \emph{B. sylvaticum} was not responsive to soil moisture. In contrast, \emph{B. sylvaticum} appears to show a more dramatic response in leaf composition estimated by N content and C:N ratio.

@@

Significant interactions between genotype and environmental parameters in a final model indicate the presence of genetic variation for plasticity (GxE) (Via and Lande 1985). For those GxE interactions where the environmental parameter is non-linear, significant GxE indicates genetic variation for the shape of reaction norms. SLA showed a significant interaction between genotype and soil moisture in \emph{B. distachyon}. In \emph{B. sylvaticum}, nitrogen content, carbon to nitrogen ratio, and root to shoot ratio all showed significant interactions between genotype and soil moisture (Table \ref{tab:models}, Figure \ref{fig:curves}). In each of these cases, the interaction between genotype and environment involved a non-linear environmental predictor, indicating not only variation for the magnitude of plasticity (i.e.~slope) but also variation in the shape of responses.

@@

We found significant non-linearity in response to a soil moisture gradient for all measured traits in at least one of the two species sampled. The best-fit function for some traits was quadratic, while other traits showed more complex responses to the environment which were best fit by a spline function. These results offer new insights with respect to the study of plant response to soil drying. By focusing on the curvature of phenotypic response as the explicitly modeled trait, we avoid contrasts of trait values expressed at arbitrary levels of soil water content which may obscure different thresholds of response among the diverse genotypes under study. SLA in \emph{B. distachyon} exhibits this pattern, as two accessions show a threshold-like response in decreasing SLA as soils become drier, and three accessions express their highest SLA at intermediate SWC. Leaf N content (on a leaf-mass basis) in \emph{B. sylvaticum} likewise shows considerable diversity of response with two accessions expressing their lowest values at intermediate SWC, one accession expressing its highest Leaf N at intermediate SWC, and one accession showing a nearly linear response along the SWC gradient. Such non-linearities in trait responses to soil moisture reinforce the need to consider the severity of extreme weather events when predicting plastic responses, especially when scaling up to investigating the ecosystem consequences of plant responses to environmental stress (Felton and Smith 2017).

@@

From a practical perspective this work highlights the value of function-valued trait approaches that may be extended to studying plant-water relations in agricultural settings. In this experiment, we investigated variation in plant responses to a gradient of soil moisture using six watering levels, which in combination with random variation in water capacity of experimental pots, produced a continuous gradient of soil moisture ranging from field capacity of the soil to strongly water-limited. In the field, multiple watering regimes in combination with random variability between plots may produce similar gradients of soil moisture. Here, water content was measured gravimetrically. New sensing technologies may be useful for quantifying soil moisture in an analogous fashion, to define soil moisture quantitatively and then apply function-valued statistical approaches to contrast trait expression among genotypes. Finally, while in this experiment we used destructive phenotyping methods to measure traits, non-destructive (and high throughput) phenotyping will enable measurement of yield or fitness data as well, to examine explicit connections between trait variation and adaptation to different degrees of soil moisture.

@@

\leavevmode\hypertarget{ref-felton2017integrating}{}%
Felton A. J., and M. D. Smith, 2017 Integrating plant ecological responses to climate extremes from individual to ecosystem levels. Philosophical Transactions of the Royal Society B: Biological Sciences 372: 20160142.
